An Analysis of the Use of Average Centipawn Loss to Detect Cheating in Chess

### 2022-11-15 Manuscript/manuscript.tex
added citations

%% preamble
\usepackage[margin = 1in]{geometry}
\usepackage{booktabs}
\usepackage{natbib}
\usepackage[colorlinks=true, citecolor=blue]{hyperref}
\usepackage{graphicx}

@@

% revisit when research is completed

@@

The first game was played “over the board” (OTB) in contrast to “virtual chess” which is hosted online. Cheating in online chess is very easy. One simply needs to run a chess engine (like Stockfish) and execute the moves from their game into the engine as they played, following up with whichever move it suggested they play next. Cheating OTB is not as simple. Players are commonly required to pass through a metal detector and are not permitted to bring phones or any other electronic devices with them (see the afore mentioned article about the famous concealed shoe computer).

The specific allegations related to how Niemann cheated OTB are unknown. Also, it would be very hard to prove cheating by simply looking at moves that are made in a single game. But what if you had access to hundreds or even thousand of games from a certain individual over a long stretch of their career?
Although cheating in online chess should be, in principle, at least as hard to detect as with cheating OTB, with the added disadvantage that you cannot enforce the disallowance of electronic devices, in reality it is quite different. Chess.com, which is used by millions of players has released an expose on their cheat detection methods, specifically with regard to Hans Niemann, who they allege has cheating in many games. The methods used include (1) comparing moves made to engine-recommended moves, (2) comparing player past performances to their historical strength, (3) comparing a player's performance to comparable peers, (4) looking at behavioral factors, and (5) reviewing time usage for finding "easy" vs. "difficult" moves.

@@

% how to make individual images that are different resolutions the same size?

\includegraphics[scale=0.3]{niemann_rating.png}

\includegraphics[scale=0.5]{carlsen_rating.png}

@@

@@

\section*{Appendix}
\addcontentsline{toc}{section}{Appendix}
\label{sec:appx}

\section*{References}
\addcontentsline{toc}{section}{References}
\label{sec:refs}

>>>>

%% preamble
\usepackage[margin = 1in]{geometry}
\usepackage{booktabs}
\usepackage{natbib}
\usepackage[colorlinks=true, citecolor=blue]{hyperref}
\usepackage{graphicx}
\usepackage{subcaption}

@@

Evolutions in chess computing have allowed for extensive libraries of past games to be analyzed again utlizing centipawn math. A recent scandal involving the alleged cheating of Chess Grandmaster Hans Niemann in the sphere of professional chess has launched a conversation about how cheaters can be detected statistically to the forefront of Academic Data Science. This paper will explore how one might detect cheating using average centipawn loss and whether these methods are reliable enough to become a standard in the chess proctors handbook.

Keywords: Centipawn math, Niemann, Carlsen, Chess

@@

The first game was played “over the board” (OTB) in contrast to “virtual chess” which is hosted online. Cheating in online chess is very easy. One simply needs to run a chess engine (like Stockfish) and execute the moves from their game into the engine as they played, following up with whichever move it suggested they play next. Cheating OTB is not as simple. Players are commonly required to pass through a metal detector and are not permitted to bring phones or any other electronic devices with them.

The specific allegations related to how Niemann cheated OTB are unknown. Also, it would be very hard to prove cheating by simply looking at moves that are made in a single game. But what if you had access to hundreds or even thousand of games from a certain individual over a long stretch of their career? Although cheating in online chess should be, in principle, at least as hard to detect as with cheating OTB, with the added disadvantage that you cannot enforce the disallowance of electronic devices, in reality it is quite different. Chess.com, which is used by millions of players has released an expose on their cheat detection methods, specifically with regard to Hans Niemann, who they allege has cheating in many games \citep{erikallebest2022}. The methods used include (1) comparing moves made to engine-recommended moves, (2) comparing player past performances to their historical strength, (3) comparing a player's performance to comparable peers, (4) looking at behavioral factors, and (5) reviewing time usage for finding "easy" vs. "difficult" moves.

@@

% how to make individual images that are different resolutions the same size?
\begin{figure}
    \begin{subfigure}{.5\textwidth}
    \includegraphics[scale=0.2]{niemann_rating.png}
    \caption{Hans Niemann's ELO ratings over time.}
    \label{fig:sfig1}
    \end{subfigure}

    \begin{subfigure}{.5\textwidth}
    \includegraphics[scale=0.45]{carlsen_rating.png}
    \caption{Magnus Carlsen's ELO ratings over time.}
    \label{fig:sfig2}
    \end{subfigure}
\label{fig:fig}
\end{figure}

@@

From the figures provided, one can see that the trends in ELO ranking over time for both Niemann and Carlsen are quite different. Although it should be noted that Niemann has had a far shorter professional career than Carlsen, one can observe that he has gone through varying periods of growth and decline whereas Carlsen has been following somewhat of a logarithmic growth pattern. Note: compare equal ranges for Carlsen and Niemann as far as ELO ranking for a more unbiased analysis, this may be inaccurate.

@@

\bibliographystyle{chicago}
\bibliography{citations}
